---
title: "Lab 3"
author: "Enoch Kim"
date: "11:59PM March 4, 2021"
output:
  word_document: default
  pdf_document: default
---


## Support Vector Machine vs. Perceptron

We recreate the data from the previous lab and visualize it:

```{r}
pacman::p_load(ggplot2)
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

Use the `e1071` package to fit an SVM model to the simple data. Use a formula to create the model, pass in the data frame, set kernel to be `linear` for the linear SVM and don't scale the covariates. Call the model object `svm_model`. Otherwise the remaining code won't work.

```{r}
pacman::p_load(e1071)
svm_model = svm(
  formula = Xy_simple$response ~., 
  data = Xy_simple, 
  kernel = "linear", 
  scale = FALSE
)
```

and then use the following code to visualize the line in purple:

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% cbind(Xy_simple$first_feature, Xy_simple$second_feature)[svm_model$index, ] # the other terms
)
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")
simple_viz_obj + simple_svm_line
```

Source the `perceptron_learning_algorithm` function from lab 2. Then run the following to fit the perceptron and plot its line in orange with the SVM's line:

```{r}

perceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w = NULL){
  
  Xinput = as.matrix(cbind(1,Xinput))
  p = ncol(Xinput)
  w = rep(0, p)
  r = nrow(Xinput)
  
  for (iter in 1 : MAX_ITER){  
    for (i in 1 : r) {
      x_i = Xinput[i, ]
      yhat_i = ifelse(sum(x_i * w) > 0, 1, 0)
      y_i = y_binary[i]
      for(j in 1:p){
        w[j] = w[j] + (y_i - yhat_i) * x_i[j]
      }
    }
  }
  w
}

w_vec_simple_per = perceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1)
)
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")
simple_viz_obj + simple_perceptron_line + simple_svm_line
```

Is this SVM line a better fit than the perceptron?

Yes, this SVM line is a better fit than the perecptron, it is marginally acceptable. 

Now write pseuocode for your own implementation of the linear support vector machine algorithm using the Vapnik objective function we discussed.

Note there are differences between this spec and the perceptron learning algorithm spec in question \#1. You should figure out a way to respect the `MAX_ITER` argument value. 


```{r}
#' Support Vector Machine 
#
#' This function implements the hinge-loss + maximum margin linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 1.
#' @return            The computed final parameter (weight) as a vector of length p + 1
linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = 0.1){
  
  #Declaring the Sum of hinge error
  # SHE = 0
  
  #  initialize a w vector
  #  w <- () (This something like this?)
  
  #Iterate through the for-loop till the MAX_ITER
  #  for x in MAX_ITER{
  
  #Iterate through each row
  #    for i in nrow(Xinput){
            
           #Adding the hinge error for each i to obtain total
  #        SHE += max{0, (1/2) - (y_binary[i]-1/2)(w*Xinput[i]-b)}
  #    }
  #  }
  #  To maximize the margin
  #  argmin{(SHE/n) + lambda(distance of w)^2}
  
  #  Returns w (the wedge/weights?) 
  #  w
}
  

```


If you are enrolled in 342W the following is extra credit but if you're enrolled in 650, the following is required. Write the actual code. You may want to take a look at the `optimx` package. You can feel free to define another function (a "private" function) in this chunk if you wish. R has a way to create public and private functions, but I believe you need to create a package to do that (beyond the scope of this course).

```{r}
#' This function implements the hinge-loss + maximum margin linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 1.
#' @return            The computed final parameter (weight) as a vector of length p + 1
#linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = 0.1){
  
#}
```

If you wrote code (the extra credit), run your function using the defaults and plot it in brown vis-a-vis the previous model's line:

```{r}
#svm_model_weights = linear_svm_learning_algorithm(Xinput, y_binary)
#my_svm_line = geom_abline(
#    intercept = svm_model_weights[1] / svm_model_weights[3],#NOTE: negative sign removed from intercept argument here
#    slope = -svm_model_weights[2] / svm_model_weights[3], 
#    color = "brown")
#simple_viz_obj  + my_svm_line
```

Is this the same as what the `e1071` implementation returned? Why or why not?

TO-DO

We now move on to simple linear modeling using the ordinary least squares algorithm.

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
```

Compute $h^*(x)$ as `h_star_x, then draw $\epsilon \sim N(0, 0.33^2)$ as `epsilon`, then compute $\y$.
```{r}
h_star_x = beta_0 + beta_1 * x 
epsilon = rnorm(n, 0, 0.33)
y = h_star_x + epsilon
```

Graph the data by running the following chunk:

```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

Does this make sense given the values of $beta_0$ and $beta_1$?

Write a function `my_simple_ols` that takes in a vector `x` and vector `y` and returns a list that contains the `b_0` (intercept), `b_1` (slope), `yhat` (the predictions), `e` (the residuals), `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you can only use the functions `sum` and `length` and other basic arithmetic operations. You should throw errors if the inputs are non-numeric or not the same length. You should also name the class of the return value `my_simple_ols_obj` by using the `class` function as a setter. No need to create ROxygen documentation here.

```{r}
my_simple_ols = function(x, y){
  n = length(y)
  
  if(n != length(x)){
    stop("x and y need to be the same length")
  }
  
  if(class(x) != 'numeric' && class(x) != 'integer'){
    stop("x needs to be numeric.")
  }
  
  if(class(y) != 'numeric' && class(y) != 'integer'){
    stop("y needs to be numeric.")
  }  
  
  if (n <= 2){
    stop("n must be more than 2")
  }
  
  x_bar = sum(x) / n
  y_bar = sum(y) / n 
  b_1 = (sum(x * y) - n * x_bar * y_bar)/(sum(x ^ 2)- n * x_bar ^ 2)
  b_0 = y_bar - b_1 * x_bar
  yhat = b_0 + b_1 * x
  e = y - yhat
  SSE = sum(e ^ 2)
  SST = sum ((y - y_bar) ^ 2)
  MSE = SSE / (n - 2)
  RMSE = sqrt(MSE)
  Rsqrt = 1 - (SSE/SST)
  
  model = list(b_0 = b_0, b_1 = b_1, yhat = yhat, e = e, SSE = SSE, SST = SST, MSE = MSE, RMSE = RMSE, Rsqrt = Rsqrt)
  
  class(model) = "my_simple_ols_obj"
  
  model
  
}
```

Verify your computations are correct for the vectors `x` and `y` from the first chunk using the `lm` function in R:

```{r}
lm_mod = lm(y ~ x)
my_simple_ols_mod = my_simple_ols(x,y)
#run the tests to ensure the function is up to spec
pacman::p_load(testthat)
expect_equal(my_simple_ols_mod$b_0, as.numeric(coef(lm_mod)[1]), tol = 1e-4)
expect_equal(my_simple_ols_mod$b_1, as.numeric(coef(lm_mod)[2]), tol = 1e-4)
expect_equal(my_simple_ols_mod$RMSE, summary(lm_mod)$sigma, tol = 1e-4)
expect_equal(my_simple_ols_mod$Rsq, summary(lm_mod)$r.squared, tol = 1e-4)
```

Verify that the average of the residuals is 0 using the `expect_equal`. Hint: use the syntax above.

```{r}
mean(my_simple_ols_mod$e)
expect_equal(mean(my_simple_ols_mod$e), 0)
```

Create the $X$ matrix for this data example. Make sure it has the correct dimension.

```{r}
X = cbind(1, x)
```

Use the `model.matrix` function to compute the matrix `X` and verify it is the same as your manual construction.

```{r}
model.matrix(~x)
```

Create a prediction method `g` that takes in a vector `x_star` and `my_simple_ols_obj`, an object of type `my_simple_ols_obj` and predicts y values for each entry in `x_star`. 

```{r}
g = function(my_simple_ols_obj, x_star){
  y_star = my_simple_ols_obj$b_0 + my_simple_ols_obj$b_1 * x_star
}
```

Use this function to verify that when predicting for the average x, you get the average y.

```{r}
expect_equal(g(my_simple_ols_mod, mean(x)), mean(y))
```


In class we spoke about error due to ignorance, misspecification error and estimation error. Show that as n grows, estimation error shrinks. Let us define an error metric that is the difference between $b_0$ and $b_1$ and $\beta_0$ and $\beta_1$. How about $h = ||b - \beta||^2$ where the quantities are now the vectors of size two. Show as n increases, this shrinks. 

```{r}
beta_0 = 3
beta_1 = -2
beta = c(beta_0, beta_1)

ns = 10 ^ (1:8)

errors_in_betas = array(NA, length(ns))

for (i in 1 : length(ns)) {
  n = ns[i]
  x = runif(n)
  h_star_x = beta_0 + beta_1 * x 
  epsilon = rnorm(n, mean = 0, sd = 0.33)
  y = h_star_x + epsilon
  
  
  mod = my_simple_ols(x,y)
  b = c(mod$b_0, mod$b_1)
  
  errors_in_betas[i] = sum((beta - b)^2)
}

log(errors_in_betas, 10)
```


We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
pacman::p_load(HistData)
```

In it, there is a dataset called `Galton`. Load it up.

```{r}
data(Galton)
```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report $n$, $p$ and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.
p is 1 and n is 928 the number of observations  

```{r}
pacman::p_load(skimr)
skim(Galton)
```

In Galton's data, I see the number of rows and columns within the data. Within the data, in the skim part. I see the parent and the child in the row, in the columns, I see the mean (parent is 68.3, child is 68.1), sd and see the percentile, 0, 25, 50, 75 and 100. $n$ is the number of observations, $p$ is the number of independent variable in this case it is height. 

Find the average height (include both parents and children in this computation).

```{r}
avg_height = mean(c(Galton$parent, Galton$child))
avg_height
```

If you were predicting the child height from the parent height and you were using null model, what would the RMSE be of this model be?

```{r}
n =  nrow(Galton) 
SST = sum((Galton$child - mean(Galton$child)) ^ 2)
sqrt(SST / (n - 1))
```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report $b_0$, $b_1$, RMSE and $R^2$. 

```{r}
mod = lm(child ~ parent, Galton) 
b_0 = coef(mod)[1]
b_1 = coef(mod)[2]
summary(mod)$sigma
summary(mod)$r.squared
```

Interpret all four quantities: $b_0$, $b_1$, RMSE and $R^2$. Use the correct units of these metrics in your answer.

$b_0$ = If x = 0, then the parent has no height that means the parent does not exist (can't have zero height, in inches) 
$b_1$ = If there is a one inch increase by the parent's height, the child average height will increases by the .64 inches.
$R^2$ = This is about 21% variance explained which is consider low to certain standards.
RMSE = This is Plus or minus 4.5 (2.23(2)) inches which makes it a 9 inches range 95% of the time

How good is this model? How well does it predict? Discuss.

This model is pretty good because it only considers one feature which is height but 9 inches can be considered a large range because
the confidence interval can be almost a foot(12 inches) off of the prediction. 

It is reasonable to assume that parents and their children have the same height? Explain why this is reasonable using basic biology and common sense.

Yes is it reasonable to assume that parents and their children have the same height because children inherit both their parent's height genes which would have the similar height as their parents. 

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of $\beta_0$ and $\beta_1$ be?

The value of beta one is 1 and value of beta zero is 0. 

Let's plot (a) the data in $\mathbb{D}$ as black dots, (b) your least squares line defined by $b_0$ and $b_1$ in blue, (c) the theoretical line $\beta_0$ and $\beta_1$ if the parent-child height equality held in red and (d) the mean height in green.

```{r}
pacman::p_load(ggplot2)
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = avg_height, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

Children of short parents became tall on average and children of tall parents became short on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?

Galton called it because if the tall parents had short children and short parents had tall parents, this led to a balance on height in the gene pool. The population by on height is regressing back to the mean. 

Why should this effect be real?

This effect should be real because as time passes by, the height's mean for the human population stays around the same point due to biology. If someone is taller or shorter than average, the population will regress to the mean. 

You now have unlocked the mystery. Why is it that when modeling with $y$ continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with $y$ continuous.

The reason why everyone calls it "regression" is because we expected the parent/child relationship to be around 1 to 1. However after running the test and viewing the results, the height average of the children seem to have regressed compared to the height average of parent. The children height average turned out to be lower than the parents height average. 

You can now clear the workspace. Create a dataset $\mathbb{D}$ which we call `Xy` such that the linear model as $R^2$ about 50\% and RMSE approximately 1.

```{r}
x = 1:5
y = x^25
Xy = data.frame(x = x, y = y)
model = lm(x~y)
summary(model)$r.squared
summary(model)$sigma
```

Create a dataset $\mathbb{D}$ which we call `Xy` such that the linear model as $R^2$ about 0\% but x, y are clearly associated.

```{r}
x = 1:400
y = x ^ 118
Xy = data.frame(x = x, y = y)
model = lm(x~y)
summary(model)$r.squared
```

Extra credit: create a dataset $\mathbb{D}$ and a model that can give you $R^2$ arbitrarily close to 1 i.e. approximately 1 - epsilon but RMSE arbitrarily high i.e. approximately M.

```{r}
epsilon = 0.01
M = 1000
#TO-DO
```


Write a function `my_ols` that takes in `X`, a matrix with with p columns representing the feature measurements for each of the n units, a vector of $n$ responses `y` and returns a list that contains the `b`, the $p+1$-sized column vector of OLS coefficients, `yhat` (the vector of $n$ predictions), `e` (the vector of $n$ residuals), `df` for degrees of freedom of the model, `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you cannot use `lm` or any other package; it must be done manually. You should throw errors if the inputs are non-numeric or not the same length. Or if `X` is not otherwise suitable. You should also name the class of the return value `my_ols` by using the `class` function as a setter. No need to create ROxygen documentation here.


```{r}
#From Office Hours

my_ols = function(X, y){
  
  n = length(y)
  
  if (!is.numeric(X) && !is.integer(X)) {
    stop("X is not numeric")
  }
  
  X = cbind(rep(1, n), X)
  
  p = ncol(X)
  
  df = ncol(X)
  
  if (n != nrow(X)){
    stop("X rows and length of y need to be the same length.")
  }
  
  if(class(y) !='numeric' && class(y) !='integer'){
    stop("y needs to be numeric.")
   }
  
  if(n<=ncol(X)+1){
    stop("n must be more than 2.")
  }
  
  y_bar = sum(y) / n
  
  b = solve(t(X) %*% X) %*% t(X) %*% y
  
  yhat =  X %*% b

  e = y - yhat
  
  SSE = (t(e) %*% e)
  SST = t(y - y_bar) %*% (y - y_bar)
  MSE = SSE / (n- (p + 1))
  RMSE = sqrt(MSE)
  RSQ = 1 - (SSE / SST)
  
  model = list(b = b, yhat = yhat, df = df, e = e, SSE = SSE, SST = SST, MSE = MSE, RMSE = RMSE, RSQ = RSQ)
  
  class(model) = "my_ols_obj"
  
  model
}
```

Verify that the OLS coefficients for the `Type` of cars in the cars dataset gives you the same results as we did in class (i.e. the ybar's within group). 

```{r}
cars = MASS::Cars93
mod = lm(Price~Type, data=cars)
my_ols(as.numeric(data.matrix(data.frame((cars$Type)))), cars$Price)
cars
```


Create a prediction method `g` that takes in a vector `x_star` and the dataset $\mathbb{D}$ i.e. `X` and `y` and returns the OLS predictions. Let `X` be a matrix with with p columns representing the feature measurements for each of the n units

```{r}
#From Office Hours

g = function(x_star, X, y){
  b = my_ols(X, y)$b
  x_star = c(1,x_star)
  x_star %*% b
}

X = model.matrix(~Type,cars)[, 2:6]
head(X)

g(X[1,], X, cars$Price)
t(c(1,X[1,])) %*% my_ols(X, cars$Price)$b
X[1,]

predict(mod, cars[1,])
```
