---
title: "Lab 8"
author: "Enoch Kim"
output: pdf_document
date: "11:59PM April 29, 2021"
---

I want to make some use of my CART package. Everyone please try to run the following:

```{r}
if (!pacman::p_isinstalled(YARF)){
  pacman::p_install_gh("kapelner/YARF/YARFJARs", ref = "dev")
  pacman::p_install_gh("kapelner/YARF/YARF", ref = "dev", force = TRUE)
}
options(java.parameters = "-Xmx4000m")
pacman::p_load(YARF)
```

For many of you it will not work. That's okay.

Throughout this part of this assignment you can use either the `tidyverse` package suite or `data.table` to answer but not base R. You can mix `data.table` with `magrittr` piping if you wish but don't go back and forth between `tbl_df`'s and `data.table` objects.

```{r}
pacman::p_load(tidyverse, magrittr, data.table)
```

We will be using the `storms` dataset from the `dplyr` package. Filter this dataset on all storms that have no missing measurements for the two diameter variables, "ts_diameter" and "hu_diameter".

```{r}
data(storms)

storms2 = storms %>%
  filter(!is.na(ts_diameter) & !is.na(hu_diameter) & ts_diameter > 0 & hu_diameter > 0)

storms2
```

From this subset, create a data frame that only has storm, observation period number for each storm (i.e., 1, 2, ..., T) and the "ts_diameter" and "hu_diameter" metrics.

```{r}

storms2 = storms2 %>% 
  select(name, ts_diameter, hu_diameter) %>%
  group_by(name) %>%
  mutate (period = row_number())

storms2
```

Create a data frame in long format with columns "diameter" for the measurement and "diameter_type" which will be categorical taking on the values "hu" or "ts".

```{r}
storms_long = pivot_longer(storms2, cols = matches("diameter"), names_to = "diameter")
storms_long
```

Using this long-formatted data frame, use a line plot to illustrate both "ts_diameter" and "hu_diameter" metrics by observation period for four random storms using a 2x2 faceting. The two diameters should appear in two different colors and there should be an appropriate legend.

```{r}
storms_sample = sample(unique(storms2$name),4)
ggplot(storms_long %>% filter(name %in% storms_sample)) + 
  geom_line(aes(x = period, y = value, col = diameter)) +
  facet_wrap(name ~. , nrow = 2)
```

In this next first part of this lab, we will be joining three datasets in an effort to make a design matrix that predicts if a bill will be paid on time. Clean up and load up the three files. Then I'll rename a few features and then we can examine the data frames:

```{r}
rm(list = ls())
pacman::p_load(tidyverse, magrittr, data.table, R.utils)
bills = fread("https://github.com/kapelner/QC_MATH_342W_Spring_2021/raw/master/labs/bills_dataset/bills.csv.bz2")
payments = fread("https://github.com/kapelner/QC_MATH_342W_Spring_2021/raw/master/labs/bills_dataset/payments.csv.bz2")
discounts = fread("https://github.com/kapelner/QC_MATH_342W_Spring_2021/raw/master/labs/bills_dataset/discounts.csv.bz2")
setnames(bills, "amount", "tot_amount")
setnames(payments, "amount", "paid_amount")
head(bills)
head(payments)
head(discounts)
bills = as_tibble(bills)
payments = as_tibble(payments)
discounts = as_tibble(discounts)
```

The unit we care about is the bill. The y metric we care about will be "paid in full" which is 1 if the company paid their total amount (we will generate this y metric later).

Since this is the response, we would like to construct the very best design matrix in order to predict y.

I will create the basic steps for you guys. First, join the three datasets in an intelligent way. You will need to examine the datasets beforehand.

```{r}
bills_with_payments = left_join(bills, payments, by = c("id" = "bill_id"))
bills_with_payments                                
bills_with_payments_with_discounts = left_join(bills_with_payments, discounts, by = c("discount_id" = "id"))
bills_with_payments_with_discounts
```

Now create the binary response metric `paid_in_full` as the last column and create the beginnings of a design matrix `bills_data`. Ensure the unit / observation is bill i.e. each row should be one bill! 

```{r}
bills_data = bills_with_payments_with_discounts%>%
  mutate(tot_amount = if_else(is.na(pct_off), tot_amount, tot_amount*(1-pct_off/100)))%>%
  group_by(id)%>%
  mutate(sum_of_payment_amount = sum(paid_amount))%>%
  mutate(paid_in_full = if_else(sum_of_payment_amount >= tot_amount, 1,0, missing =0  ))%>%
  slice(1) %>%
  ungroup()
table(bills_data$paid_in_full, useNA = "always")
```

How should you add features from transformations (called "featurization")? What data type(s) should they be? Make some features below if you think of any useful ones. Name the columns appropriately so another data scientist can easily understand what information is in your variables.

```{r}
pacman::p_load("lubridate")
bills_data = bills_data %>%
  select(-id, -id.y, -num_days, -transaction_date, -pct_off, -days_until_discount, -sum_of_payment_amount, -paid_amount) %>%
  mutate(num_days_to_pay = as.integer(ymd(due_date) - ymd(invoice_date))) %>%
  select(-due_date, -invoice_date) %>%
  mutate(discount_id = as.factor(discount_id)) %>%
  group_by(customer_id) %>%
  mutate(bill_num = row_number()) %>%
  ungroup() %>%
  select(-customer_id, -discount_id) %>%
  relocate(paid_in_full, .after = last_col())

bills_data
```

Now let's do this exercise. Let's retain 25% of our data for test.

```{r}
K = 4
test_indices = sample(1 : nrow(bills_data), round(nrow(bills_data) / K))
train_indices = setdiff(1 : nrow(bills_data), test_indices)
bills_data_test = bills_data[test_indices, ]
bills_data_train = bills_data[train_indices, ]
```

Now try to build a classification tree model for `paid_in_full` with the features (use the `Xy` parameter in `YARF`). If you cannot get `YARF` to install, use the package `rpart` (the standard R tree package) instead. You will need to install it and read through some documentation to find the correct syntax.

Warning: this data is highly anonymized and there is likely zero signal! So don't expect to get predictive accuracy. The value of the exercise is in the practice. I think this exercise (with the joining exercise above) may be one of the most useful exercises in the entire semester.

```{r}
#install.packages('rpart')
pacman::p_load(rpart)
mod1 = rpart(paid_in_full ~., data = bills_data_train, method = "class")
mod1
```

For those of you who installed `YARF`, what are the number of nodes and depth of the tree? 

```{r}
nrow(mod1$frame) ##number of nodes
```

For those of you who installed `YARF`, print out an image of the tree.

```{r}
plot(mod1, uniform=TRUE)
text(mod1, use.n=TRUE, all=TRUE, cex=.8)
```

Predict on the test set and compute a confusion matrix.

```{r}
yhat = predict(mod1, bills_data_test, type = c("class"), na.action = na.pass)
oos_conf_table = table(bills_data_test$paid_in_full, yhat)
oos_conf_table
```

Report the following error metrics: misclassifcation error, precision, recall, F1, FDR, FOR.

```{r}
#Levels
n = sum(oos_conf_table)
fp = oos_conf_table[1, 2]
fn = oos_conf_table[2, 1]
tp = oos_conf_table[2, 2]
tn = oos_conf_table[1, 1]
num_pred_pos = sum(oos_conf_table[, 2])
num_pred_neg = sum(oos_conf_table[, 1])
num_pos = sum(oos_conf_table[2, ])
num_neg = sum(oos_conf_table[1, ])

misclassifcation_error = (fn + fp) / n
cat("misclassifcation_error", round(misclassifcation_error * 100, 2), "%\n")
precision = tp / num_pred_pos
cat("precision", round(precision * 100, 2), "%\n")
recall = tp / num_pos
cat("recall", round(recall * 100, 2), "%\n")
false_discovery_rate = 1 - precision
cat("false_discovery_rate", round(false_discovery_rate * 100, 2), "%\n")
false_omission_rate = fn / num_pred_neg
cat("false_omission_rate", round(false_omission_rate * 100, 2), "%\n")
F1 = (2 * tp)/(2 * tp + fp + fn)
cat("F1", round(F1 * 100, 2), "%\n")
```

Is this a good model? (yes/no and explain).

Depends, on what someone thinks a the term "good". In this case, the best case scenario would be that FP stays lower compare to FN.
When it comes to prediction, you rather have someone who is not going to pay, end up paying then having someone who was predicted to pay but not end up paying because someone paying but ends up not paying, can really damage the business. This model is "alright"

There are probability asymmetric costs to the two types of errors. Assign the costs below and calculate oos total cost.

```{r}
C_fp = 69
C_fn = 1
cost = C_fp * fp + C_fn * fn
cost
```

We now wish to do asymmetric cost classification. Fit a logistic regression model to this data.

```{r}
logistic_mod = glm(paid_in_full ~ ., bills_data_train, family = binomial(link = "logit"))
#p_hats_train = predict(logistic_mod, bills_data_train, type = "response")
```

Use the function from class to calculate all the error metrics for the values of the probability threshold being 0.001, 0.002, ..., 0.999 in a data frame.

```{r}
compute_metrics_prob_classifier = function(p_hats, y_true, res = 0.001){
  #we first make the grid of all prob thresholds
  p_thresholds = seq(0 + res, 1 - res, by = res) #values of 0 or 1 are trivial
  
  #now we create a matrix which will house all of our results
  performance_metrics = matrix(NA, nrow = length(p_thresholds), ncol = 12)
  colnames(performance_metrics) = c(
    "p_th",
    "TN",
    "FP",
    "FN",
    "TP",
    "miscl_err",
    "precision",
    "recall",
    "FDR",
    "FPR",
    "FOR",
    "miss_rate"
  )
  
  #now we iterate through each p_th and calculate all metrics about the classifier and save
  n = length(y_true)
  for (i in 1 : length(p_thresholds)){
    p_th = p_thresholds[i]
    y_hats = factor(ifelse(p_hats >= p_th, 1, 0))
    confusion_table = table(
      factor(y_true, levels = c(0, 1)),
      factor(y_hats, levels = c(0, 1))
    )
      
    fp = confusion_table[1, 2]
    fn = confusion_table[2, 1]
    tp = confusion_table[2, 2]
    tn = confusion_table[1, 1]
    npp = sum(confusion_table[, 2])
    npn = sum(confusion_table[, 1])
    np = sum(confusion_table[2, ])
    nn = sum(confusion_table[1, ])
  
    performance_metrics[i, ] = c(
      p_th,
      tn,
      fp,
      fn,
      tp,
      (fp + fn) / n,
      tp / npp, #precision
      tp / np,  #recall
      fp / npp, #false discovery rate (FDR)
      fp / nn,  #false positive rate (FPR)
      fn / npn, #false omission rate (FOR)
      fn / np   #miss rate
    )
  }
  
  #finally return the matrix
  performance_metrics
}

p_hats_train = predict(logistic_mod, bills_data_train, type = "response")
p_hats_test = predict(logistic_mod, bills_data_test, type = "response")
y_true = bills_data_train$paid_in_full
y_true_2 = bills_data_test$paid_in_full

metric_prob_classifier_in_sample = compute_metrics_prob_classifier(p_hats_train, y_true) %>% data.table
metric_prob_classifier_in_sample_tibble = as_tibble(metric_prob_classifier_in_sample)
metric_prob_classifier_out_sample = compute_metrics_prob_classifier(p_hats_test, y_true_2) %>% data.table
metric_prob_classifier_out_sample_tibble = as_tibble(metric_prob_classifier_out_sample)
```

Calculate the column `total_cost` and append it to this data frame.

```{r}
C_fp = 69
C_fn = 1

metric_prob_classifier_in_sample_tibble = metric_prob_classifier_in_sample_tibble %>%
  mutate(total_cost = C_fp * FP + C_fn * FN)
metric_prob_classifier_in_sample_tibble

metric_prob_classifier_out_sample_tibble = metric_prob_classifier_out_sample_tibble %>%
  mutate(total_cost = C_fp * FP + C_fn * FN)
metric_prob_classifier_out_sample_tibble

```

Which is the winning probability threshold value and the total cost at that threshold?

```{r}
winning_prob_threshold_insample = which.min(metric_prob_classifier_in_sample_tibble$total_cost)
winning_prob_threshold_insample_metrics = metric_prob_classifier_in_sample_tibble[winning_prob_threshold_insample,]

cat("The winning probability threshold value in-sample is:", min(winning_prob_threshold_insample_metrics$total_cost))

winning_prob_threshold_outsample = which.min(metric_prob_classifier_out_sample_tibble$total_cost)
winning_prob_threshold_outsample_metrics = metric_prob_classifier_out_sample_tibble[winning_prob_threshold_outsample, ]

cat("\n \nThe winning probability threshold value out-sample is:", min(winning_prob_threshold_outsample_metrics$total_cost))
```

Plot an ROC curve and interpret.

```{r}
pacman::p_load(ggplot2)
metrics_in_and_out_performance = rbind(
    cbind(metric_prob_classifier_in_sample_tibble, data.table(sample = "in")),
    cbind(metric_prob_classifier_out_sample_tibble, data.table(sample = "out"))
)
ggplot(metrics_in_and_out_performance) +
  geom_line(aes(x = FPR, y = recall)) +
  geom_abline(intercept = 0, slope = 1, col = "Red") + 
  coord_fixed() + xlim(0, 1) + ylim(0, 1)
```

ROC stands for the "receiver operator curve". The predictive power can be calculated by the area under the curve and the Red line can be used to compare probability of the models of estimation. 

Calculate AUC and interpret.

```{r}
pacman::p_load(pracma)
auc_in_sample = -trapz(metric_prob_classifier_in_sample_tibble$FPR, metric_prob_classifier_in_sample_tibble$recall)
cat("AUC in-sample: ", auc_in_sample)

auc_oos = -trapz(metric_prob_classifier_out_sample_tibble$FPR, metric_prob_classifier_out_sample_tibble$recall)
cat("\n\nAUC out-sample: ", auc_oos)
```

The AUC in-sample is 0.5960914 and the AUC out-sample is 0.5960528. Due AUC being greater than 0.5, this model has predictive power. 

Plot a DET curve and interpret.

```{r}
ggplot(metrics_in_and_out_performance) +
  geom_line(aes(x = FDR, y = miss_rate, col = sample)) +
  coord_fixed() + xlim(0, 1) + ylim(0, 1)
```

In this plot, the DET curve shows a trade off between FDR and FOR, I don't really know how to interpret this but when FDR is around 0.42 and the miss_rate is around .98, that seems to be the start of the optimal trade off points/plot. 







